Competing Constructions for German Measure Noun Phrases: from Usage Data to Experimental Validation

### 2017-04-04 Latex/article.tex
minor fixes

The distribution of the case patterns in the NAC (case identity between the measure noun and the kind noun) and in the PGC (genitive on the kind noun, regardless of the case of the measure noun) depending on the structure of the kind NP are summarised in Table~\ref{tab:constructions}.
I call the narrow apposition construction with a bare kind noun the \NACb, the partitive genitive with a determiner in the kind noun phrase the \PGCd.
I call the alternating variants with an adjective but no determiner in the kind noun phrase \NACa, and \PGCa, respectively.
This paper is about the middle column of Table~\ref{tab:constructions}, \ie, the syntactic configuration in which two different case patterns are acceptable.

@@

Finally, (\ref{it:intro:other}) suggests that there are alternative ways of expressing similar quantificational meanings.
In the variationist tradition, which is strongly influenced by Labovian sociolinguistics, the \textit{principle of accountability} would dictate that proper studies should examine a variationist \textit{variable}, \ie\ all different \textit{ways of saying the same thing} (\citealp{Labov1966}, \citealp{Labov1969}, for an overview see \citealp{Tagliamonte2012}).
In the case at hand, the variable might be something like \textit{measuring quantities of substances and collections}.
I argue that it is fully justified to focus narrowly on \NACa\ and \PGCa\ with their well-defined morph-syntactic properties, mostly because the alternative constructions are not used in the same range of contexts.
Two major alternatives might be considered.
First of all, the analytic (pseudo-)partitive with \textit{von} `of' is only available as an alternative to the PGC if the kind noun phrase contains a (definite or indefinite) determiner as in (\ref{ex:analyticalpartitive}).

@@

This morpho-syntactic ambiguity means that the \NACa\ is in fact a \NACb in disguise, and the \PGCa\ is a \PGCd in disguise.
This should ideally be reflected in the following basic selection effect:
It is known from frameworks like Collexeme Analysis \citep{GriesStefanowitsch2004} that lemmas are attracted with different strengths by competing constructions.
If the \NACa\ is highly similar to the \NACb\ and the \PGCa\ is highly similar to the \PGCd, the probability with which individual noun lemmas appear in the alternating constructions should be predictable at least partly from the relative frequency with which the same lemmas are used in the non-alternating constructions.
Nouns which occur proportionally more often in the \PGCd\ should favour the \PGCa, and nouns which occur proportionally more often in the \NACb\ should favour the \NACa.
This basic attraction effect will be quantified in the corpus study Section~\ref{sec:corpusstudies}.

@@

The full set of manually annotated variables for the main sample is given in Table~\ref{tab:variables}, and I briefly discuss it now.%
\footnote{All numeric variables were also z-transformed (\ie, centered to the mean and rescaled such that they have a standard deviation of $1$) to facilitate their interpretation in the regression models reported in the next section.}
Notice that \textit{Construction} is the response variable (or `dependent variable') with the values \textit{PGCa} and \textit{NACa}.

>>>>

The distribution of the case patterns in the NAC (case identity between the measure noun and the kind noun) and in the PGC (genitive on the kind noun, regardless of the case of the measure noun) depending on the structure of the kind NP are summarised in Table~\ref{tab:constructions}.
I call the narrow apposition construction with a bare kind noun the \NACb, the partitive genitive with a determiner in the kind noun phrase the \PGCd.
I call the alternating variants with an adjective but no determiner in the kind noun phrase \NACa, and \PGCa, respectively.
This paper is about the middle column of Table~\ref{tab:constructions}, \ie\ the syntactic configuration in which two different case patterns are acceptable.

@@

Finally, (\ref{it:intro:other}) suggests that there are alternative ways of expressing similar quantificational meanings.
In the variationist tradition, which is strongly influenced by Labovian sociolinguistics, the \textit{principle of accountability} would dictate that proper studies should examine a variationist \textit{variable}, \ie\ all different \textit{ways of saying the same thing} (\citealp{Labov1966}, \citealp{Labov1969}, for an overview see \citealp{Tagliamonte2012}).
In the case at hand, the variable might be something like \textit{measuring quantities of substances and collections}.
I argue that it is fully justified to focus narrowly on \NACa\ and \PGCa\ with their well-defined morpho-syntactic properties, mostly because the alternative constructions are not used in the same range of contexts.
Two major alternatives might be considered.
First of all, the analytic (pseudo-)partitive with \textit{von} `of' is only available as an alternative to the PGC if the kind noun phrase contains a (definite or indefinite) determiner as in (\ref{ex:analyticalpartitive}).

@@

This morpho-syntactic ambiguity means that the \NACa\ is in fact a \NACb in disguise, and the \PGCa\ is a \PGCd in disguise.
This should ideally be reflected in the following basic selection effect:
It is known from frameworks like Collexeme Analysis \citep{GriesStefanowitsch2004} that lemmas are attracted with different strengths by competing constructions.
If the \NACa\ is highly similar to the \NACb\ and the \PGCa\ is highly similar to the \PGCd, the probability with which individual noun lemmas appear in the alternating constructions should be predictable at least partly from the relative frequency with which the same lemmas are used in the non-alternating constructions (see \citealp[246--249]{Levshina2016} for a similar idea in a different context).
Nouns which occur proportionally more often in the \PGCd\ should favour the \PGCa, and nouns which occur proportionally more often in the \NACb\ should favour the \NACa.
This basic attraction effect will be quantified in the corpus study Section~\ref{sec:corpusstudies}.

@@

The full set of manually annotated variables for the main sample is given in Table~\ref{tab:variables}, and I briefly discuss it now.%
\footnote{All numeric variables were also z-transformed (\ie\ centered to the mean and rescaled such that they have a standard deviation of $1$) to facilitate their interpretation in the regression models reported in the next section.}
Notice that \textit{Construction} is the response variable (or `dependent variable') with the values \textit{PGCa} and \textit{NACa}.
